Critical assessment of Linear-Response Excitations in Molecular system\\ or \\ On the Localization of the Linear-Response excitations in Open Systems

### 2018-01-31 LRDraft.tex
update

Thanks to the usage of the basis of transition, we have been able to express the response density as a linear superposition of excitations $\op E$. The perturbing field 
determines the relative contribution of each excitation through the weighting coefficients in the numerator of \eqref{RhopExcitationDef1}.
The fundamental question raised in the present paper, namely the ....reliable assessment of the results of the linear response can thus be reformulated in this language.

@@

We introduce the \emph{excited states} defined via the linear combination of the basis set orthogonal to $\dmnot$ with the
coefficients that appears in \eqref{XYcoeffDef1}, that is:
\begin{align}\lb{ExcitonicStatesDef1}
\ket{\phi^E_p} &\equiv \sint \dd \alpha X_{E}^{p \alpha} \ket{\psi(\alpha)} &
\tbra{\phi^E_p} &\equiv \sint \dd \alpha {\tilde X_{E}^{p \alpha}} \bra{\psi(\alpha)} \;, \nn\\
\bra{\chi^E_p} &\equiv \sint \dd \alpha  Y_{E}^{p \alpha} \bra{\psi(\alpha)} &
\tket{\chi^E_p} &\equiv \sint \dd \alpha {\tilde Y}_{E}^{p\alpha} \ket{\psi(\alpha)} \;.
\end{align}
Thanks to these definitions we can express the excitation operators as follows:
\be\lb{ExctitationOperatorsDef2}
\op E = \sum_p \left( \ket{\phi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^E_p}\right)\;, \qq
\op {\tilde E} = \sum_p \left( \ket{\psi_p} \tbra{\phi^E_p} + \tket{\chi^E_p}\bra{\psi_p} \right)\;.
\ee
Starting from these equation we can derive the excitation expansion of the fluctuation states. Indeed an inspection of formula \eqref{rhoPrimeFluctuationStateDef1}
show that fluctuation states can be extracted from the response density operator as follows:
\be
\ket{f_p(\omega)} = \dm'(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(-\omega)} = \bra{\psi_p}\dm'(\omega) \;, \nn
\ee
so plugging equations \eqref{ExctitationOperatorsDef2} in the formula \eqref{RhopExcitationDef1} and performing these operations we obtain:
\be\lb{FluctStateExcitationDef1}
\ket{f_p(\omega)} = \sint\dd E \frac{\trace{\op{\tilde E}\opskew{\Phi}(\omega)}}{\bbomega-E} \ket{\phi^E_p} \;, \qq
\bra{f_p(-\omega)} = \sint\dd E \bra{\chi^E_p}\frac{\trace{\op{\tilde E}\opskew{\Phi}(\omega)}}{\bbomega-E} 
\ee

%\subsubsection{Sternheimer equation for excitonic states. Localization argument revisited}
\subsubsection{Sternheimer equation for excited states. Localization argument revisited}

% We introduce the \emph{excitonic states} defined via the linear combination of the basis set orthogonal to $\dmnot$ with the
% coefficients that appears in \eqref{XYcoeffDef1}, that is:
% \begin{align}\lb{ExcitonicStatesDef1}
% \ket{\phi^E_p} &\equiv \sint \dd \alpha X_{E}^{p \alpha} \ket{\psi(\alpha)} &
% \tbra{\phi^E_p} &\equiv \sint \dd \alpha {\tilde X_{E}^{p \alpha}} \bra{\psi(\alpha)} \;, \nn\\
% \bra{\chi^E_p} &\equiv \sint \dd \alpha  Y_{E}^{p \alpha} \bra{\psi(\alpha)} &
% \tket{\chi^E_p} &\equiv \sint \dd \alpha {\tilde Y}_{E}^{p\alpha} \ket{\psi(\alpha)} \;.
% \end{align}
% Thanks to these definitions we can express the excitation operators as follows:
% \be\lb{ExctitationOperatorsDef2}
% \op E = \sum_p \left( \ket{\phi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^E_p}\right)\;, \qq
% \op {\tilde E} = \sum_p \left( \ket{\psi_p} \tbra{\phi^E_p} + \tket{\chi^E_p}\bra{\psi_p} \right)\;.
% \ee
The operator eigenvalue equations \eqref{opEeigenvalueDef1} and \eqref{optildeEeigenvalueDef1} can be translated into a Schroedinger-like equations
for the excitonic states.

@@

\noindent In this case the $\optr{L}_{q\beta}^{p\alpha}$ matrix has only diagonal non-vanishing elements, given by: 
\be
\trace{\decay{\alpha}{p}\Liouv\excite{q}{\beta}} = H_{0\alpha\beta}\delta_{pq} - \eps_p\delta_{pq}\delta(\alpha-\beta) \;, \qq
\trace{\excite{\alpha}{p}\Liouv\decay{q}{\beta}} = \eps_p\delta_{pq}\delta(\alpha-\beta) - H_{0\beta\alpha}\delta_{pq} \;, 
\ee
where $H_{0\alpha\beta} = \bra{\psi(\alpha)}\hnot\ket{\psi(\beta)}$. The eigenvalue conditions \eqref{LiouvMatrixEigenvectorDef1} imply
that the $Y$ and $\tilde Y$ coefficients introduced in \eqref{XYcoeffDef1} vanish so that both $\tket{\chi^E_p}$ and $\bra{\chi^E_p}$ do not
contribute to the excitation operators \eqref{ExctitationOperatorsDef2}. The eigenvalue operator equation for $\op E$ reduces to:
\be
\sum_q\left(\hnot - \eps_q\right)\ket{\phi^E_q}\bra{\psi_q} = E \sum_q \ket{\phi^E_q}\bra{\psi_q} \;,
\ee
and projecting on $\ket{\psi_q}$ provides the eigenvalue condition for the excitonic state $\ket{\phi^E_p}$:
\be
\left(\hnot - \eps_p\right)\ket{\phi^E_p} = E \ket{\phi^E_p} \;.
\ee
In the same way, by using the eigenvalue condition for $\op{\tilde E}$, we can derive an analogous equation for $\tbra{\phi^E_p}$. To achieve this
result we need the expression of the left action of the (unperturbed) Liouville superoperator defined in \eqref{L0sopDef1}, that can be derived by
observing that:
\be\lb{leftActionLouv0Def1}
\trace{\op O' (\Liouvnot \op O)} = \trace{(\op O' \Liouvnot) \op O } =
\trace{\commutator{\op O'}{\op H_0} \op O} \quad \Rightarrow \quad \op O \Liouvnot = 
- \Liouvnot \op O \;.
\ee
So, the eigenvalue operator equation for $\op{\tilde E}$ provides:
\be
\sum_q  \ket{\psi_q} \tbra{\phi^E_q} \left( \hnot - \eps_q \right) =
E \sum_q  \ket{\psi_q} \tbra{\phi^E_q} \;,
\ee
so that:
\be
\tbra{\phi^E_p} \left( \hnot - \eps_p \right) = E \tbra{\phi^E_p} \;.
\ee
So we find that the excitonic states are the left and right eigenvectors of $\hnot$ (that actually coincide since we are considering
hermitian Hamiltonians). The energy eigenvalue is shifted by $\eps_p$, that is the energy of the occupied states involved in the 
excitation.

@@

To explicitly write the generic eigenvalue equations for the excitonic states it is useful to introduce some further formalism to express the 
self-consistent part of the perturbing Hamiltonian \eqref{Vptspacedef1}. We start by expanding $\op V'$ in the coordinate representation, that is:
\be
\op V'(t)= 
\int \dd \br \dd \br' \dd \r \dd \r' \bra{\br'}\frac{\delta\hnot}{\delta \rho_0(\r,\r')}\ket{\br}
\trace{\dm'(t) \ket{\r} \bra{\r'}} \ketbra{\br'}{\br}\;, \nn
\ee
then we introduce the \emph{coupling kernel}, defined as:
\be\lb{couplingKernelDef1}
U\left[\op O; \op O'\right] \equiv  \int \dd \r \dd \r' \trace{\op O \frac{\delta \hnot }{\delta \rho_0(\r,\r')}
} \trace{\op O' \ket{\r} \bra{\r'}} \;,
\ee
using this quantity we can express $\op V'$ as:
\be\lb{Vptspacedef2}
\op V'(t)= 
\int \dd \br \dd \br' U\left[\ketbra{\br}{\br'};\dm'(t)\right] \ketbra{\br'}{\br}\;,
\ee
and consequently the formal action of the coupling superoperator \eqref{couplingSopDef1} on a generic operator $\op O$ can be expressed as: 
\be\lb{couplingSopDef2}
\coupl\op O = 
\int \dd \r \dd \r' \commutator{\ketbra{\r'}{\r}}{\dmnot} U\left[\ketbra{\r}{\r'};\op O\right] \;.
\ee
Using this formalism we can define the left action of the coupling superoperator, in analogous way of what done in \eqref{leftActionLouv0Def1}, 
that is:
\begin{align}\lb{leftActionCouplingDef1}
\trace{\op O' (\coupl \op O)} &= \int \dd\r\dd\r'\trace{\ketbra{\r'}{\r}\commutator{\dmnot}{\op O'}}
U\left[\ketbra{\r}{\r'};\op O\right]  = \nn \\ &= 
U\left[\commutator{\dmnot}{\op O'};\op O\right] \doteq \trace{(\op O' \coupl) \op O } \;,
\end{align}
where the second equality follows from the linearity of the coupling kernel. The right action of the coupling superoperator consistent with this result is:
\be\lb{leftActionCouplingDef2}
(\op O \coupl) = \int \dd\r\dd\r'
U\left[\commutator{\dmnot}{\op O};\ketbra{\r'}{\r}\right]\ketbra{\r}{\r'} \;.
\ee
This formalism allows us to provide a compact expression for the action of the Liouville superoperator on the excitation operators. For the left action we have:
\begin{align}
\Liouv\op E &= \sum_p\left(\hnot-\eps_p\right)\ketbra{\phi^E_p}{\psi_p} - \sum_p\ketbra{\psi_p}{\chi^E_p}\left(\hnot-\eps_p\right) + \nn \\
&+ \sum_p \sint \dd\alpha  U\left[\decay{\alpha}{p}; \op E\right] \ketbra{\psi(\alpha)}{\psi_p}
- \sum_p \sint \dd\alpha U\left[\excite{p}{\alpha}; \op E\right] \ketbra{\psi_p}{\psi(\alpha)} = E\op E \;, \nn
\end{align}
here the action of the coupling superoperator on $\op E$ has been computed imposing its expansion in the transition basis and computing the coefficients with formula 
\eqref{leftActionCouplingDef1}. Projecting out the previous equation by multiplication with an occupied eigenstates from both the left and right provides the equations:
\begin{align}\lb{PhipChipEqDef1}
\left(\hnot-\eps_p\right)\ket{\phi^E_p} +\sint \dd\alpha  U\left[\decay{\alpha}{p}; \op E\right] \ket{\psi(\alpha)} &= E\ket{\phi^E_p} \;,\nn \\
\bra{\chi^E_p}\left(\hnot-\eps_p\right) +\sint \dd\alpha \bra{\psi(\alpha)} U\left[\excite{p}{\alpha}; \op E\right] &= -E\bra{\chi^E_p} \;.
\end{align}
We observe that the coupling kernel behaves as a (non-local) potential that depends on both the $X_{E}^{p \alpha}$ and $Y_{E}^{p \alpha}$ coefficients of \eqref{ExcitonicStatesDef1}
thus introducing a coupling between the $\phi$ and $\chi$ states.

>>>>

Thanks to the usage of the basis of transition, we have been able to express the response density as a linear superposition of excitation operators. Now 
we are interested in analyzing this restatement of the linear response under the perspective of the main topic of the paper, namely the critical assessment 
of the results provided by the linear response formalism in a given computational setup. Our interests are mainly twofold. First, to analyze the localization 
properties of the excitations, in a analogous fashion of what done in section \ref{LocalizationFluctuationStates} for the fluctuation states. Second, to determine 
the precise nature of the formal sum over excitations written in \eqref{RhopExcitationDef1}, in order to be able to discriminate between the discrete and the 
continuum sector of the excitation spectrum.

@@

%Thanks to the usage of the basis of transition, we have been able to express the response density as a linear superposition of excitations $\op E$. Now
We push this formalism a little bit forward and derive the expansion of the fluctuation states \eqref{fluctuationStateDef1} in terms of the \emph{excited states} of the 
system . Firstly, we introduce excited states, defined via the linear combination of the elements of the $\alpha$-basis with the coefficients that appears in 
\eqref{XYcoeffDef1}, that is:
\begin{align}\lb{ExcitedStatesDef1}
\ket{\phi^E_p} &\equiv \sint \dd \alpha X_{E}^{p \alpha} \ket{\psi(\alpha)} &
\tbra{\phi^E_p} &\equiv \sint \dd \alpha {\tilde X_{E}^{p \alpha}} \bra{\psi(\alpha)} \;, \nn\\
\bra{\chi^E_p} &\equiv \sint \dd \alpha  Y_{E}^{p \alpha} \bra{\psi(\alpha)} &
\tket{\chi^E_p} &\equiv \sint \dd \alpha {\tilde Y}_{E}^{p\alpha} \ket{\psi(\alpha)} \;,
\end{align}
due to these definitions we can express excitation operators as follows:
\be\lb{ExcitationOperatorsDef2}
\op E = \sum_p \left( \ket{\phi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^E_p}\right)\;, \qq
\op {\tilde E} = \sum_p \left( \ket{\psi_p} \tbra{\phi^E_p} + \tket{\chi^E_p}\bra{\psi_p} \right)\;.
\ee
Formulas \eqref{ExcitationOperatorsDef2} provides a clear representation of the physical meaning of excited states \eqref{ExcitedStatesDef1}, as can be seen by applying
the excitation operators on the occupied eigenstates of the systems:
\be
\op E\ket{\psi_p} = \ket{\phi^E_p} \;, \quad \bra{\psi_p} \op E = \bra{\chi^E_p} \;, \quad
\op {\tilde E}\ket{\psi_p} = \tket{\chi^E_p} \;, \quad \bra{\psi_p} \op {\tilde E} = \tbra{\phi^E_p} \;.
\nn
\ee
The excited states expansion of fluctuation states can be easily derived.  Indeed an inspection of formula \eqref{rhoPrimeFluctuationStateDef1} show that fluctuation 
states can be extracted from the response density operator as follows:
\be
\ket{f_p(\omega)} = \dm'(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(-\omega)} = \bra{\psi_p}\dm'(\omega) \;, \nn
\ee
so plugging the expression \eqref{ExcitationOperatorsDef2} inside \eqref{RhopExcitationDef1} and applying the response density to a occupied state provides:
\be\lb{FluctStateExcitationDef1}
\ket{f_p(\omega)} = \sint\dd E \frac{\trace{\op{\tilde E}\opskew{\Phi}(\omega)}}{\bbomega-E} \ket{\phi^E_p} \;, \qq
\bra{f_p(-\omega)} = \sint\dd E \bra{\chi^E_p}\frac{\trace{\op{\tilde E}\opskew{\Phi}(\omega)}}{\bbomega-E} \;.
\ee
Fluctuation states emerge as a linear superposition of excited states. The $\omega$-dependent coefficients codify the effect of the perturbing field in determining the 
relative contribution of each addend through the weighting coefficients in the numerators of \eqref{FluctStateExcitationDef1}.

We perform a particular choice of the perturbing field:
\be
\op\Phi = \commutator{\op E}{\dmnot} \; \longrightarrow \; \opskew{\Phi} = 
\commutator{\commutator{\op E}{\dmnot}}{\dmnot} = \op E
\ee

\subsubsection{Sternheimer equation for excited states. Localization argument revisited}

The operator eigenvalue equations \eqref{opEeigenvalueDef1} and \eqref{optildeEeigenvalueDef1} can be translated into a Schroedinger-like equations
for the excitonic states.

@@

\noindent In this case the $\optr{L}_{q\beta}^{p\alpha}$ matrix has only diagonal non-vanishing elements, given by: 
\be
\trace{\decay{\alpha}{p}\Liouv\excite{q}{\beta}} = H_{0\alpha\beta}\delta_{pq} - \eps_p\delta_{pq}\delta(\alpha-\beta) \;, \qq
\trace{\excite{\alpha}{p}\Liouv\decay{q}{\beta}} = \eps_p\delta_{pq}\delta(\alpha-\beta) - H_{0\beta\alpha}\delta_{pq} \;, 
\ee
where $H_{0\alpha\beta} = \bra{\psi(\alpha)}\hnot\ket{\psi(\beta)}$. The eigenvalue conditions \eqref{LiouvMatrixEigenvectorDef1} imply
that the $Y$ and $\tilde Y$ coefficients introduced in \eqref{XYcoeffDef1} vanish so that both $\tket{\chi^E_p}$ and $\bra{\chi^E_p}$ do not
contribute to the excitation operators \eqref{ExcitationOperatorsDef2}. The eigenvalue operator equation for $\op E$ reduces to:
\be
\sum_q\left(\hnot - \eps_q\right)\ket{\phi^E_q}\bra{\psi_q} = E \sum_q \ket{\phi^E_q}\bra{\psi_q} \;,
\ee
and projecting on $\ket{\psi_q}$ provides the eigenvalue condition for the excitonic state $\ket{\phi^E_p}$:
\be
\left(\hnot - \eps_p\right)\ket{\phi^E_p} = E \ket{\phi^E_p} \;.
\ee
In the same way, by using the eigenvalue condition for $\op{\tilde E}$, we can derive an analogous equation for $\tbra{\phi^E_p}$. To achieve this
result we need the expression of the left action of the (unperturbed) Liouville superoperator defined in \eqref{L0sopDef1}, that can be derived by
observing that:
\be\lb{leftActionLouv0Def1}
\trace{\op O' (\Liouvnot \op O)} = \trace{(\op O' \Liouvnot) \op O } =
\trace{\commutator{\op O'}{\op H_0} \op O} \quad \Rightarrow \quad \op O \Liouvnot = 
- \Liouvnot \op O \;.
\ee
So, the eigenvalue operator equation for $\op{\tilde E}$ provides:
\be
\sum_q  \ket{\psi_q} \tbra{\phi^E_q} \left( \hnot - \eps_q \right) =
E \sum_q  \ket{\psi_q} \tbra{\phi^E_q} \;,
\ee
so that:
\be
\tbra{\phi^E_p} \left( \hnot - \eps_p \right) = E \tbra{\phi^E_p} \;.
\ee
So we find that the excitonic states are the left and right eigenvectors of $\hnot$ (that actually coincide since we are considering
hermitian Hamiltonians). The energy eigenvalue is shifted by $\eps_p$, that is the energy of the occupied states involved in the 
excitation.

@@

To explicitly write the generic eigenvalue equations for the excitonic states it is useful to introduce some further formalism to express the 
self-consistent part of the perturbing Hamiltonian \eqref{Vptspacedef1}. We start by expanding $\op V'$ in the coordinate representation, that is:
\be
\op V'(t)= 
\int \dd \br \dd \br' \dd \r \dd \r' \bra{\br'}\frac{\delta\hnot}{\delta \rho_0(\r,\r')}\ket{\br}
\trace{\dm'(t) \ket{\r} \bra{\r'}} \ketbra{\br'}{\br}\;, \nn
\ee
then we introduce the \emph{coupling kernel}, defined as:
\be\lb{couplingKernelDef1}
U\left[\op O; \op O'\right] \equiv  \int \dd \r \dd \r' \trace{\op O \frac{\delta \hnot }{\delta \rho_0(\r,\r')}
} \trace{\op O' \ket{\r} \bra{\r'}} \;,
\ee
using this quantity we can express $\op V'$ as:
\be\lb{Vptspacedef2}
\op V'(t)= 
\int \dd \br \dd \br' U\left[\ketbra{\br}{\br'};\dm'(t)\right] \ketbra{\br'}{\br}\;,
\ee
and consequently the formal action of the coupling superoperator \eqref{couplingSopDef1} on a generic operator $\op O$ can be expressed as: 
\be\lb{couplingSopDef2}
\coupl\op O = 
\int \dd \r \dd \r' \commutator{\ketbra{\r'}{\r}}{\dmnot} U\left[\ketbra{\r}{\r'};\op O\right] \;.
\ee
Using this formalism we can define the left action of the coupling superoperator, in analogous way of what done in \eqref{leftActionLouv0Def1}, 
that is:
\begin{align}\lb{leftActionCouplingDef1}
\trace{\op O' (\coupl \op O)} &= \int \dd\r\dd\r'\trace{\ketbra{\r'}{\r}\commutator{\dmnot}{\op O'}}
U\left[\ketbra{\r}{\r'};\op O\right]  = \nn \\ &= 
U\left[\commutator{\dmnot}{\op O'};\op O\right] \doteq \trace{(\op O' \coupl) \op O } \;,
\end{align}
where the second equality follows from the linearity of the coupling kernel. The right action of the coupling superoperator consistent with this result is:
\be\lb{leftActionCouplingDef2}
(\op O \coupl) = \int \dd\r\dd\r'
U\left[\commutator{\dmnot}{\op O};\ketbra{\r'}{\r}\right]\ketbra{\r}{\r'} \;.
\ee
This formalism allows us to provide a compact expression for the action of the Liouville superoperator on the excitation operators. For the left action we have:
\begin{align}
\Liouv\op E &= \sum_p\left(\hnot-\eps_p\right)\ketbra{\phi^E_p}{\psi_p} - \sum_p\ketbra{\psi_p}{\chi^E_p}\left(\hnot-\eps_p\right) + \nn \\
&+ \sum_p \sint \dd\alpha  U\left[\decay{\alpha}{p}; \op E\right] \ketbra{\psi(\alpha)}{\psi_p}
- \sum_p \sint \dd\alpha U\left[\excite{p}{\alpha}; \op E\right] \ketbra{\psi_p}{\psi(\alpha)} = E\op E \;, \nn
\end{align}
here the action of the coupling superoperator on $\op E$ has been computed imposing its expansion in the transition basis and computing the coefficients with formula 
\eqref{leftActionCouplingDef1}. Projecting out the previous equation by multiplication with an occupied eigenstates from both the left and right provides the equations:
\begin{align}\lb{PhipChipEqDef1}
\left(\hnot-\eps_p\right)\ket{\phi^E_p} +\sint \dd\alpha  U\left[\decay{\alpha}{p}; \op E\right] \ket{\psi(\alpha)} &= E\ket{\phi^E_p} \;,\nn \\
\bra{\chi^E_p}\left(\hnot-\eps_p\right) +\sint \dd\alpha \bra{\psi(\alpha)} U\left[\excite{p}{\alpha}; \op E\right] &= -E\bra{\chi^E_p} \;.
\end{align}
We observe that the coupling kernel behaves as a (non-local) potential that depends on both the $X_{E}^{p \alpha}$ and $Y_{E}^{p \alpha}$ coefficients of \eqref{ExcitedStatesDef1}
thus introducing a coupling between the $\phi$ and $\chi$ states.
